---
title: "session2"
author: "Kevin Wang and Garth Tarr"
date: "05/11/2019"
output: html_document
---


# Learning outcomes of this session

+ Summaring a `data.frame`
+ Making basic plots using `ggplot2`: tile plot and scatter plots
+ Joining two datasets together (`left_join`)
+ `pivot_longer` and `pivot_wider` to reshape data for plotting purposes



# Loading packages

```{r}
library(tidyverse)
library(skimr)
```


# Loading cleaned sample data

```{r}
session1_data = read_csv(file = "data/clean_sample_data.csv")
```

# Skimming your data
```{r}
skimr::skim(session1_data)
```


# Mutating data
```{r}
clean_sample_data = session1_data %>% 
  dplyr::mutate(dose = as.character(dose),
                time = as.character(time))
```



# Making summary statistics 

## Counting number of samples
```{r}
clean_sample_data %>% 
  group_by(strain) %>% 
  count()

clean_sample_data %>% 
  group_by(strain) %>% 
  summarise(n())
```

## Counting with more variables
```{r}
clean_sample_data %>% 
  group_by(strain, dose) %>% 
  count()

clean_sample_data %>% 
  group_by(strain, dose, time) %>% 
  count()
```


# Making simple plots 
```{r}
count_data = clean_sample_data %>% 
  group_by(strain, dose, time) %>% 
  count() %>% 
  ungroup()


count_data %>% 
  ggplot(aes(x = dose, y = time, fill = n)) +
  geom_tile()

count_data %>% 
  ggplot(aes(x = dose, y = time, fill = n)) +
  geom_tile() +
  facet_grid(~strain)
```


## (Extension: ordering using factors)
```{r}
class(session1_data$dose)
unique(session1_data$dose)
class(count_data$dose)
unique(count_data$dose)

dose_order = sort(unique(session1_data$dose))

clean_count_data = count_data %>% 
  dplyr::mutate(dose = factor(dose, dose_order))

clean_count_data %>% 
  ggplot(aes(x = dose, y = time, fill = n, label = n)) +
  geom_tile() +
  geom_text() +
  facet_grid(~strain)
```



```{r, include=FALSE}
time_order = sort(unique(session1_data$time))

clean_count_data = count_data %>% 
  dplyr::mutate(dose = factor(dose, dose_order),
                time = factor(time, time_order))

clean_count_data %>% 
  ggplot(aes(x = dose, y = time, fill = n, label = n)) +
  geom_tile() +
  geom_text() +
  facet_grid(~strain)
```


# Reading in gene expression data (top 10 rows)
```{r}
raw_ge_data = read_delim(file = "data/GSE43251_NanoString_non-normalized.txt", delim = "\t", n_max = 5)
# raw_ge_data = read_tsv(file = "data/GSE43251_NanoString_non-normalized.txt", n_max = 5)

raw_ge_data = raw_ge_data %>% 
  dplyr::select(-1)
# raw_ge_data = raw_ge_data %>% 
#   dplyr::select(-GB_ACC)
```




# left_join sample data
```{r}
ge_sample_data = data.frame(
  sample = colnames(raw_ge_data)[-1],
  ge_data_available = "yes"
)

joined_sample_data = left_join(clean_sample_data, ge_sample_data, by = "sample")

joined_sample_data
```


# `tidyr::pivot_longer` the gene expression data

```{r}
ge_data_long = raw_ge_data %>% 
  tidyr::pivot_longer(
    cols = -GENE_SYMBOL,
    names_to = "sample", 
    values_to = "gene_expression")

ge_data_long
```


```{r}
long_sample_ge_data = clean_sample_data %>% 
  left_join(ge_data_long, by = "sample") %>% 
  dplyr::filter(!is.na(GENE_SYMBOL))
```



# `tidyr::pivot_wider` the gene expression data

```{r}
wide_sample_ge_data = long_sample_ge_data %>% 
  tidyr::pivot_wider(names_from = "GENE_SYMBOL",
                     values_from = "gene_expression")
```


```{r}
write_csv(long_sample_ge_data, path = "data/long_sample_ge_data.csv")
write_csv(wide_sample_ge_data, path = "data/wide_sample_ge_data.csv")
```


# Session Info
```{r}
sessionInfo()
```

